Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-19 paper-filmat.tex
Tackled a couple of XXX's

XXX mention connection to realizability

\subsection{Exploring the \effective topos}
\label{sect:eff-examples}
Due to its computational nature, truth in the \effective topos is quite
different from truth in the standard topos. The following examples are XXX

\begin{tabular}{lll}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially so) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is the zero function or not. & \cmark{} (trivially so) & \xmark \\
  Any function $\NN \to \NN$ is computable by a Turing machine. & \xmark & \cmark{} (trivially so) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  \bottomrule
\end{tabular}
% XXX formatting

@@

% XXX external ...
Its external meaning when interpreted in the \effective topos is that there exists
a Turing machine which, when given a natural number~$n$ as input, terminates with a
prime number~$p > n$ as output. Such a Turing machine exists, hence the
statement is true in the \effective topos.

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is the zero function or not.''} More
formally, the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
The preceding examples give the impression that what is true in the
\effective topos is simply a subset of what is true in the standard topos. The
example of this subsection shows that the relation between the two toposes is more nuanced.

@@

The law of trichotomy, stating that any real number is either negative, zero or
positive, generally fails in~$\Sh(X)$. By the Kripke--Joyal semantics, the external
% XXX external
meaning of the internal statement~``$\forall a:\RR\_ a<0 \vee a=0 \vee a>0$'' is that for any continuous function~$a : U
\to \RR$ defined on any open~$U$ of~$X$, there is an open covering~$U =
\bigcup_i U_i$ such that on each member~$U_i$ of this covering, the function~$a$ is
either everywhere negative on~$U_i$, everywhere zero on~$U_i$ or everywhere
positive on~$U_i$. But this statement is, for most base spaces~$X$, false.
Figure~\ref{fig:trichotomy}(c) shows a counterexample.

@@

* XXX Syntactical vs. semantical interpretation
* XXX phone call analogy
* XXX quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
* XXX allowing a switch of perspective
* XXX applications in mathematical practice
* XXX int. logic as common denominator
* XXX arbitrariness of the "standard axioms"
* XXX importance of having an adapted language
* XXX Source https://plato.stanford.edu/entries/category-theory/ for relevant literature
* XXX the internal language as a generalized modal operator
* XXX use the phrase "very rich and very deep" (or similarly) about set theory
* XXX Sven(?)
* XXX semantic view vs. syntactic view
* XXX Kripke models
* XXX more pointers to topos literature
* XXX Reference for the effective topos
* XXX A topos is more than its semantics
* XXX Reference: Moerdijk
* XXX Check references everywhere (give good pointers)
* XXX Format rules in a better way
* XXX toposes are not small (Figure 1)
* XXX hypercomputation or supercomputation?

>>>>

% FUTURE (in separate notes about the effective topos): Relationship to
% realizability

\subsection{Exploring the \effective topos}
\label{sect:eff-examples}
Due to its computational nature, truth in the \effective topos is quite
different from truth in the standard topos. This section explores the following
examples:

\bigskip
\begin{center}
\begin{tabular}{lll}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is constantly zero or not. & \cmark{} (trivially) & \xmark \\
  Any function $\NN \to \NN$ is computable. & \xmark & \cmark{} (trivially) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Markov's principle holds. & \cmark{} (trivially) & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  \bottomrule
\end{tabular}
\end{center}

@@

Its external meaning when interpreted in the \effective topos is that there exists
a Turing machine~$M$ which, when given a natural number~$n$ as input, terminates with a
prime number~$p > n$ as output. Such a Turing machine exists, hence the
statement is true in the \effective topos.\footnote{More precisely, the
machine~$M$ should also output the description of a Turing machine which
witnesses that~$p$ is prime. However, the statement~``$p$ is prime''
is~$\neg\neg$-stable (even decidable), and for those statements witnesses are
redundant.}

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is constantly zero or not.''} More
formally, the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is computable.''}
The preceding examples give the impression that what is true in the
\effective topos is simply a subset of what is true in the standard topos. The
example of this subsection shows that the relation between the two toposes is more nuanced.

@@

The law of trichotomy, stating that any real number is either negative, zero or
positive, generally fails in~$\Sh(X)$. By the Kripke--Joyal semantics, the external
meaning of the internal statement~``$\forall a:\RR\_ a<0 \vee a=0 \vee a>0$'' is that for any continuous function~$a : U
\to \RR$ defined on any open~$U$ of~$X$, there is an open covering~$U =
\bigcup_i U_i$ such that on each member~$U_i$ of this covering, the function~$a$ is
either everywhere negative on~$U_i$, everywhere zero on~$U_i$ or everywhere
positive on~$U_i$. But this statement is, for most base spaces~$X$, false.
Figure~\ref{fig:trichotomy}(c) shows a counterexample.

@@

* XXX Syntactical vs. semantical interpretation
* XXX phone call analogy
* XXX quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
* XXX allowing a switch of perspective
* XXX applications in mathematical practice
* XXX int. logic as common denominator
* XXX arbitrariness of the "standard axioms"
* XXX importance of having an adapted language
* XXX Source https://plato.stanford.edu/entries/category-theory/ for relevant literature
* XXX the internal language as a generalized modal operator
* XXX use the phrase "very rich and very deep" (or similarly) about set theory
* XXX Sven(?)
* XXX semantic view vs. syntactic view
* XXX Kripke models
* XXX more pointers to topos literature
* XXX Reference for the effective topos
* XXX A topos is more than its semantics
* XXX Reference: Moerdijk
* XXX Check references everywhere (give good pointers)
* XXX Format rules in a better way
* XXX toposes are not small (Figure 1)
* XXX hypercomputation or supercomputation?
* XXX internal/external
